Making Sequential Programs Deterministic

### 2019-04-05 java-design.tex
Move and edit defaults section

\subsection{Defaulting rules}\label{defaulting}

\TheDeterminismChecker applies a default qualifier at each unannotated type
use (except uses of type variables, which already stand for a type that was
defaulted at the instantiation site).
This does not change the expressivity of the type system; it merely makes
the system more convenient to use by reducing programmer effort and code clutter.

The default annotation for most unannotated types is \<@Det>, because
programmers generally expect their programs to behave the same when re-run
on the same inputs.

@@

\noindent
This choice type-checks if the method body does not make calls to any
interfaces that require \<@Det> arguments or produce \<@NonDet> results.

\begin{itemize}
\item
  If a formal parameter/return type is an array, its component type defaults to
  \<@PolyDet>.
\item
  As an exception, the return type of a method with no unannotated or
  \<@PolyDet> formal parameters (including the receiver) defaults to
  \<@Det>.
\end{itemize}

These choices permit all possible arguments to an unannotated method.

In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
method parameters and return types including constructors.
The rationale for this design choice was to not make \theDeterminismChecker too restrictive by making everything \<Det>.

\begin{verbatim}
@PolyDet int foo(@PolyDet int x) { ... }

void bar(@NonDet int random, @Det int value) {
    @NonDet int x = foo(random);      // no error
    @Det int y = foo(value);          // no error
    System.out.println(foo(random));  // error
}
\end{verbatim}
In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. \TheDeterminismChecker flags an error 
only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
breaks our determinism guarantees).

>>>>

\subsection{Defaulting}\label{defaulting}

\TheDeterminismChecker applies a default qualifier at each unqualified Java
basetype (except uses of type variables, which already stand for a type that was
defaulted at the instantiation site).
This does not change the expressivity of the type system; it merely makes
the system more convenient to use by reducing programmer effort and code clutter.
Defaulted type qualifiers are not trusted:  they are type-checked just as
explicitly-written ones are.

@@

\noindent
This choice type-checks if the method body does not make calls to any
interfaces that require \<@Det> arguments or produce \<@NonDet> results.
Otherwise, the programmer must write an explicit type qualifier.

If a formal parameter or return type is a collection, its element type
defaults to \<@PolyDet>.  No other choice would result in a valid type
after instantiation.

As an exception to the above rules about return types, if a method's formal
parameters (including the receiver) are all \<@Det> --- that is, there are
no unannotated or \<@PolyDet> formal parameters --- then an unannotated
return type defaults to \<@Det>.  This is particularly useful for methods
that take no formal parameters.  A type like $\forallt{\kappa} ()
\rightarrow \kappa int$ does not make sense, becauses there is no way to
instantiate the type argument $\kappa$.  Treating the type as $()
\rightarrow \<Det> int$ permits just as many uses.

The default annotation for local variables is \<@NonDet>, which is the top
qualifier in the hierarchy.  \Cref{sec:dataflow-java} explains why.

The default annotation for other unannotated types is \<@Det>, because
programmers generally expect their programs to behave the same when re-run
on the same inputs.
